Some examples in extended logic counting classes

### 2016-12-28 LICS/counting.tex
Nueva versión del apendice.

Then, what functional complexity classes can be captured with fragments of $\qso$?
For answering this question, it is reasonable to start with $\shp$ who is one of the most well-known functional complexity class~\cite{arora2009computational}. 
Since $\shp$ has a strong similarity with $\np$, one could expect a ``Fagin-like'' Theorem~\cite{fagin1974generalized} for this class. 
Actually, in~\cite{SalujaST95} it was shown that the class $\sfo$ captures $\shp$.
In our setting, the class $\sfo$ is contained in $\eqso(\fo)$ which also captures $\shp$ as expected. 
\begin{proposition} \label{eqso-fo-cap}
	$\eqso(\fo)$ captures $\shp$ over ordered structures.
\end{proposition}
Recall that (see Section~\ref{sec:previous}) that every function class $\# \LL$ is contained in $\eqso(\LL)$. Thus, by~\cite{SalujaST95} it directly follows that every $\shp$-function can be defined in $\eqso(\LL)$. The other direction is also direct and follows by showing that $\shp$ is closed under first- and second-order sum.
Moreover, these closure properties also hold if we change the boolean core $\fo$ to any polynomial-time logic like $\lfp$. This gives the following corollary.
\begin{corollary} \label{eqso-lfp-cap}
	$\eqso(\lfp)$ captures $\shp$ over ordered structures.
\end{corollary}
By following the same approach as~\cite{SalujaST95}, Compton et al.~\cite{ComptonG96} show that $\seso$ captures $\spp$. As one could expect, if we parametrize $\eqso$ with $\eso$, we can also capture~$\spp$.
\begin{proposition} \label{eqso-eso-cap}
	$\eqso(\eso)$ captures $\spp$ over ordered structures.
\end{proposition}
The previous result connecting fragments from $\# \LL$ with $\shp$ and $\spp$ is the further that we can go by using SO-formulas with free variables. 
So, what happen with other counting complexity classes like $\fp$ and $\fpspace$?
By trying to search for an $\# \LL$ class for capturing $\fp$ one can easily see that this is pointless.
Indeed, by using free SO-variables one will easily capture $\shp$-hard functions (see Section~\ref{sec:syntactic}) and by only using free FO-variables one cannot capture exponential functions definable in $\fp$ (e.g. $2^{n}$).
Therefore, it is impossible to capture classes like $\fp$ or $\fpspace$ if we follow previous approaches like~\cite{SalujaST95}.

Instead, if we move from $\eqso$ we have other options for counting like first- and second-order products. Actually, if we combine $\qfo$ with $\lfp$ this is exactly what we need to capture $\fp$, the class of function that can be computed in~$\ptime$.
\begin{theorem} \label{captfp}
	$\qfo(\lfp)$ captures $\fp$ over ordered structures.
\end{theorem}
To capture $\fp$ one first shows that any formula in $\qfo(\lfp)$ can be evaluated in $\ptime$. 
Indeed, $\lfp$ is a polynomial time logic~\cite{Immerman82,vardi1982complexity} and the sum and product quantifiers can be computed in $\fp$. 
For the other direction, one has to use $\lfp$ to simulate the run of the $\fp$-machine and use the quantitative quantifiers to reconstruct the binary number built in the tape. 
It is important to notice that the alternation between sum and products quantifiers is crucial for capturing $\fp$.
That is, without these operators is difficult to capture $\fp$.

@@

From the proof of the previous theorem a natural question follows: what happen if we use first-order quantitative quantifiers and $\pfp$?
In~\cite{Ladner89}, Ladner also proposes the class $\nfpspace$ of all functions computed by $\pspace$-machines whose output is polynomially bounded. 
Interestingly, if we restrict to FO-quantitative quantifiers we can also capture this class.
\begin{corollary} \label{qfo-pfp-cap}
	$\qfo(\pfp)$ captures $\nfpspace$ over ordered structures.
\end{corollary}

>>>>

Then, what functional complexity classes can be captured with fragments of $\qso$?
For answering this question, it is reasonable to start with $\shp$ who is one of the most well-known functional complexity class~\cite{arora2009computational}. 
Since $\shp$ has a strong similarity with $\np$, one could expect a ``Fagin-like'' Theorem~\cite{fagin1974generalized} for this class. 
Actually, in~\cite{SalujaST95} it was shown that the class $\sfo$ captures $\shp$.
In our setting, the class $\sfo$ is contained in $\eqso(\fo)$ which also captures $\shp$ as expected. 
\begin{proposition} \label{prop:capture-shP}
	$\eqso(\fo)$ captures $\shp$ over ordered structures.
\end{proposition}
Recall that (see Section~\ref{sec:previous}) that every function class $\# \LL$ is contained in $\eqso(\LL)$. Thus, by~\cite{SalujaST95} it directly follows that every $\shp$-function can be defined in $\eqso(\LL)$. The other direction is also direct and follows by showing that $\shp$ is closed under first- and second-order sum.
Moreover, these closure properties also hold if we change the boolean core $\fo$ to any polynomial-time logic like $\lfp$. This gives the following corollary.
\begin{corollary} \label{eqso-lfp-cap}
	$\eqso(\lfp)$ captures $\shp$ over ordered structures.
\end{corollary}
By following the same approach as~\cite{SalujaST95}, Compton et al.~\cite{ComptonG96} show that $\seso$ captures $\spp$. As one could expect, if we parametrize $\eqso$ with $\eso$, we can also capture~$\spp$.
\begin{proposition} \label{eqso-eso-cap}
	$\eqso(\eso)$ captures $\spp$ over ordered structures.
\end{proposition}
The previous result connecting fragments from $\# \LL$ with $\shp$ and $\spp$ is the further that we can go by using SO-formulas with free variables. 
So, what happen with other counting complexity classes like $\fp$ and $\fpspace$?
By trying to search for an $\# \LL$ class for capturing $\fp$ one can easily see that this is pointless.
Indeed, by using free SO-variables one will easily capture $\shp$-hard functions (see Section~\ref{sec:syntactic}) and by only using free FO-variables one cannot capture exponential functions definable in $\fp$ (e.g. $2^{n}$).
Therefore, it is impossible to capture classes like $\fp$ or $\fpspace$ if we follow previous approaches like~\cite{SalujaST95}.

Instead, if we move from $\eqso$ we have other options for counting like first- and second-order products. Actually, if we combine $\qfo$ with $\lfp$ this is exactly what we need to capture $\fp$, the class of function that can be computed in~$\ptime$.
\begin{theorem} \label{theo:capture-fp}
	$\qfo(\lfp)$ captures $\fp$ over ordered structures.
\end{theorem}
To capture $\fp$ one first shows that any formula in $\qfo(\lfp)$ can be evaluated in $\ptime$. 
Indeed, $\lfp$ is a polynomial time logic~\cite{Immerman82,vardi1982complexity} and the sum and product quantifiers can be computed in $\fp$. 
For the other direction, one has to use $\lfp$ to simulate the run of the $\fp$-machine and use the quantitative quantifiers to reconstruct the binary number built in the tape. 
It is important to notice that the alternation between sum and products quantifiers is crucial for capturing $\fp$.
That is, without these operators is difficult to capture $\fp$.

@@

From the proof of the previous theorem a natural question follows: what happen if we use first-order quantitative quantifiers and $\pfp$?
In~\cite{Ladner89}, Ladner also proposes the class $\nfpspace$ of all functions computed by $\pspace$-machines whose output is polynomially bounded. 
Interestingly, if we restrict to FO-quantitative quantifiers we can also capture this class.
\begin{corollary} \label{cor:capture-fpspace-poly}
	$\qfo(\pfp)$ captures $\nfpspace$ over ordered structures.
\end{corollary}
